---
title: "Email Marketing: What's the impact of subject line copy and time-sent on open and click-through rates? DRAFT"
author: "Steffen Hartleib"
output: html_document
---

### Overview:

We spent 9 months testing and tweaking subjectlines of daily deal emails. We managed to keep open rates almost constant even though we trippled the number of emails we sent per day. We improved the conversion rate by 20%. Now that we have data on over 500 different marketing emails I wanted to see if there are any patterns. For example:   
Are there any words that tend to produce better open and click through rates?   
Does the weekday or the time of day matter?  
How important is the name of the artist?   
  
```{r packages, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }
options(scipen=10000)

library('dplyr')
library('dplyr')
library('stringr')
library('ggplot2')
library('reshape2')
```

### Read Data

```{r read_data, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }
setwd("/Users/steffen/Google Drive/ds_projects/subject_lines")

emails <- read.csv("data/email_campaigns_report.csv", header = TRUE, stringsAsFactors = FALSE)

products  = read.csv("data/products_export.csv", header = TRUE, stringsAsFactors = FALSE)
products$UPC = str_replace_all(products$UPC, "'","")
products$Artist = trimws(products$Artist, which = c("right"))

lines_items = read.csv("data/line_items_final_export.csv", header = TRUE, stringsAsFactors = FALSE)
pm_lines = filter(lines_items, Store_ID %in% c('1', '5693042'))
pm_line_art = inner_join(pm_lines, products, by = c("Lineitem_sku" = "UPC" ))
```


##### Create a list of artist names who's products previously sold, along with total revenue per artist. We'll use Total Revenue as a proxy for popularity. Assuming that more popular artists get higher open rates.

```{r artist_list, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }
artist_rank = select(pm_line_art, Artist, Lineitem_quantity, Lineitem_price)%>%
              mutate(Rev = Lineitem_quantity * Lineitem_price)%>%
              mutate(Artist = str_replace(Artist,"The ",""))%>%
              mutate(Artist = ifelse(Artist == "Bob Dylan", "Dylan", Artist))%>%
              group_by(Artist)%>%
              summarize(TL_Rev = sum(Rev))%>%
              arrange(desc(TL_Rev))
           
artist_rank <-mutate(artist_rank,Artist = str_replace(artist_rank$Artist,"The ",""))%>%
              mutate(Artist = ifelse(Artist == "Bob Dylan", "Dylan", Artist))%>%
              mutate(Artist = ifelse(Artist == "Rolling Stones", "Stones", Artist))%>%
              mutate(Artist = ifelse(Artist == "David Bowie", "Bowie", Artist))%>%
              mutate(Artist = ifelse(Artist == "Led Zeppelin", "Zeppelin", Artist))%>%
              mutate(Artist = ifelse(Artist == "Ozzy Osbourne", "Ozzy", Artist))%>%
              mutate(Artist = ifelse(Artist == "ACDC", "AC/DC", Artist))%>%
              mutate(Artist = ifelse(Artist == "Bruce Springsteen", "Springsteen", Artist))%>%
              mutate(Artist = ifelse(Artist == "Black Sabbath", "Sabbath", Artist))%>%
              mutate(Artist = ifelse(Artist == "Jimi Hendrix", "Hendrix", Artist))%>%
              mutate(Artist = ifelse(Artist == "Paul McCartney", "McCartney", Artist))%>%
              mutate(Artist = ifelse(Artist == "John Coltrane", "Coltrane", Artist))%>%
              mutate(Artist = ifelse(Artist == "Elvis Presley", "Elvis", Artist))%>%
              filter(Artist != "The")%>%
              filter(TL_Rev > 0)%>%
              filter(Artist != "popmarket")
head(artist_rank)
```


### Feature Engineering

##### Extract, dates, times, and key terms from subject lines, and create variables:

```{r features, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }
camp <- filter(emails, sends > 100000)%>%
        select(id,last_run_date, name, subject, sends, opens, clicks)%>%
        mutate(Date = as.Date(last_run_date))

camp$date  <- as.Date(substr(camp$last_run_date,1,10))
camp$day <- as.factor(weekdays(camp$date))
camp$time <- substr(camp$last_run_date,11,25)
camp$hour <- as.numeric(substr(camp$last_run_date,12,13))
camp$or <- round(camp$opens/camp$sends,2)
camp$ctr <- round(camp$clicks/camp$sends,4)
camp$length = nchar(camp$subject)

camp$perc_off <-  str_extract(camp$subject,'..%')
camp$perc_off <- as.numeric(str_replace(camp$perc_off,'%',""))
camp$perc_off[is.na(camp$perc_off) == TRUE] <- 0

camp$amt_off <-  str_extract(camp$subject,'\\$...')
camp$amt_off <-  as.numeric(str_replace(camp$amt_off,'\\$',""))
camp$amt_off[is.na(camp$amt_off) == TRUE] <- 0

camp$off = as.numeric(grepl("Off|off|OFF",camp$subject))
camp$dis = as.numeric(grepl("Discount",camp$subject))
camp$sav = as.numeric(grepl("[S,s]avings|[S,s]ave",camp$subject))
camp$lim_ed = as.numeric(grepl('Limited Edition',camp$subject))
camp$vinyl = as.numeric(grepl('Vinyl|LP',camp$subject))
camp$box_set = as.numeric(grepl('Box Set',camp$subject))
camp$deluxe = as.numeric(grepl('Deluxe',camp$subject))
camp$exclusive = as.numeric(grepl('Exclusive',camp$subject))
camp$new = as.numeric(grepl('New',camp$subject))
camp$blow = as.numeric(grepl('Blowout',camp$subject))
camp$sale = as.numeric(grepl('Sale',camp$subject))
camp$steal = as.numeric(grepl('[S,s]teal', camp$subject))
camp$garage = as.numeric(grepl('[G,g]arage', camp$subject))
camp$surprise = as.numeric(grepl('[S,s]urprise', camp$subject))
str(camp)
```


##### Looping through artist names and subject lines to add up the Total Revenue (= value) of the artists mentioned in the subject line.

```{r artist_loop, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }
# list of artists names
artists = artist_rank$Artist

# initialize output variables 
output = data.frame(1,2)
names(output) <-  c('artists', "value")
row = 0

# loop through subjects and check each for artist names 
for(s in camp$subject) {
  a_list = c()
  a_value = 0
  row = row +1
    for(a in artists) {
      
      if(grepl(a,s) == TRUE){
        a_list = paste0(a_list," ",a)
        value = artist_rank$TL_Rev[artist_rank$Artist == a]
        a_value = a_value + value
   }
   else{
     a_list = a_list
     a_value = a_value
   }
    }
  
  if(is.null(a_list) == TRUE) { 
    a_list <- "" 
  }

line <-  c(a_list,a_value)
output <-  rbind(output, line)

}


# join artist name and value columns to 'camp' data frame 

artist_value <- output[-1,]
camp = cbind(camp,artist_value)
camp$value = round(as.numeric(camp$value),0)
```
##### For every subject line we now have  
1. a field with the artist name(s) mentioned in the subject line
2. a field with the sum of the values associated with those artists

```{r artist_and_value, echo = FALSE, eval = TRUE}
head(artist_value)
```

***

### Let's Plot

#### Open Rates versus Dollar Amount Off, Percentage Off, and Artist Value:

```{r plots, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE }

camp_melt_or_cont <- select(camp, or, amt_off, perc_off, value)%>%
  melt(id = "or")


plot_or_cont <- ggplot(camp_melt_or_cont, aes(x = value, y = or)) + 
  facet_wrap(~variable, scales = "free") + 
  geom_point() + geom_smooth(method = "lm")

plot_or_cont
```

#### => Amount Off and Percentage Off seem to have a slightly negative impact or open rates - didn't expect that.



###  Click Through Rates versus Dollar Amount Off, Percent Off, and Artist Value.
##### (CTR = clicks/sends)

```{r plots_3, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE }

camp_melt_ctr_cont <- select(camp, ctr, amt_off, perc_off, value)%>%
  melt(id = "ctr")


plot_or_cont <- ggplot(camp_melt_ctr_cont, aes(x = value, y = ctr)) + 
  facet_wrap(~variable, scales = "free") + 
  geom_point() + geom_smooth(method = "lm")

plot_or_cont
```

##### => Percentage Off reduces the open rate slightly, but increases the click through rate by a good amount. So the net effect looks positive.

***

### Open Rate versus specific terms, such as Discount, Limited, Steal ..., etc.

```{r plots_2, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE }

camp_melt_or_cat <- select(camp, or, off, dis, sav, lim_ed, vinyl, box_set,
                           deluxe, exclusive, new, blow, sale, steal, garage,
                           surprise)%>%
  melt(id = 'or')

camp_melt_or_cat <- select(camp, or, garage, surprise, off, dis, sav, lim_ed, vinyl,
                           box_set, deluxe, exclusive, new, blow, sale, steal, garage,
                           surprise)%>%
  mutate(garage = factor(garage))%>%
  melt(id = 'or')

plot_or_cat <- ggplot(camp_melt_or_cat, aes(x = value, y = or)) + 
  facet_wrap(~variable, scales = "free") + 
  geom_boxplot()

plot_or_cat
```


##### => terms that convey value such as Garage [sale], Surprise, Off etc.) are associated with a lower open rate. Again, that surprised me.  Savings is the exception.
##### => "Box set", "Vinyl", "Exclusive", and "Deluxe" are also asscociated with lower open rates. Again, that's a bit surprising, because these formats tend to sell well, I'd expected open rates to be higher for subject lines with these terms.


### Click Through Rate versus specific terms, such as Discount, Limited, Steal ... etc.

```{r plots_4, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE }
camp_melt_ctr_cat <- select(camp, ctr, off, dis, sav, lim_ed, vinyl, box_set,
                            deluxe, exclusive, new, blow, sale, steal, garage,
                            surprise)%>%
  melt(id = 'ctr')

camp_melt_ctr_cat <- select(camp, ctr, garage, surprise, off, dis, sav, lim_ed, vinyl,
                            box_set, deluxe, exclusive, new, blow, sale, steal, garage,
                            surprise)%>%
  mutate(garage = factor(garage))%>%
  melt(id = 'ctr')

plot_ctr_cat <- ggplot(camp_melt_ctr_cat, aes(x = value, y = ctr)) + 
  facet_wrap(~variable, scales = "free") + 
  geom_boxplot()

plot_ctr_cat

```

##### => "Surprise, "Savings", "Steal', and "Blow [out]" have a positive impact on Click Through Rates, even though they had a slightly negative impact on Open Rates. But the ooveral effect of including these terms appears to be positive.



### Open and click throuh rates for days of the week

```{r plots_5, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE }

plot_or_day <- ggplot(camp, aes(day, or)) + geom_boxplot() + ggtitle("Open Rate versus Day of the Week")
plot_ctr_day <- ggplot(camp, aes(day, ctr)) + geom_boxplot() + ggtitle("Click through Rate versus Day of the Week")

plot_or_day
plot_ctr_day 


# open and click throuh rates for time of day

camp$hour = as.factor(camp$hour)
plot_or_day <- ggplot(camp, aes(factor(hour), or)) + geom_boxplot()
plot_or_day
plot_ctr_day <- ggplot(camp, aes(factor(hour), ctr)) + geom_boxplot()
plot_ctr_day
plot_camp_per_hour <- ggplot(camp, aes(hour)) + geom_histogram(stat = 'count')
```

##### => Saturday and 6 pm appears to be the best time for click throuh rates - probably not statistically significant though.


### Open and click through rates over time



```{r plots_6, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE }

# open and click throuh rates for days of the week
plot_or_time <- ggplot(camp, aes(date, or)) + geom_point() + geom_smooth(method = 'lm')
plot_or_time

plot_ctr_time <- ggplot(camp, aes(date, ctr)) + geom_point() + geom_smooth(method = 'lm')
plot_ctr_time
```




```{r, eval = TRUE, echo = TRUE, message = FALSE}
# Linear models 


# Setting Open Rate (or) as the dependent variable (looking at click through later)

# select variables
camp_vars_or <- select(camp, or, hour, day, length, perc_off, amt_off, off, dis, sav, lim_ed, vinyl,
                    box_set, deluxe, exclusive, new, blow, sale, steal, garage, surprise, value)

# model all  variables
mod_or_all = lm(or ~., camp_vars_or)
summary(mod_or_all)

# compile resultsd
mod_or_all_R_squared <- summary(mod_or_all)$r.squared 
mod_or_all_F_value <-   summary(mod_or_all)$f[1]
mod_or_all_res <- c(mod_or_all_R_squared,mod_or_all_F_value)


# model with significant variables only
mod_or_2 = lm(or ~ day + off + dis + vinyl + sale + surprise +  value, camp_vars_or)
summary(mod_or_2)

# compling results
mod_or_2_R_squared <- summary(mod_or_2)$r.squared 
mod_or_2_F_value <-   summary(mod_or_2)$f[1]
mod_or_2_res <- c(mod_or_2_R_squared, mod_or_2_F_value)


# model without the 'sale' variable
mod_or_3 = lm(or ~ day + off + dis + vinyl + surprise + value, camp_vars_or)
summary(mod_or_3)

# compiling results
mod_or_3_R_squared <- summary(mod_or_3)$r.squared 
mod_or_3_F_value <-   summary(mod_or_3)$f[1]
mod_or_3_res <- c(mod_or_3_R_squared, mod_or_3_F_value)

# compiling results from all 3 open rate - models
mod_or_res <- rbind(mod_or_all_res, mod_or_2_res, mod_or_3_res)
mod_or_res



# Click Through Rate (ctr) ~ all

camp_vars_ctr <- select(camp, ctr, hour, day, length, perc_off, amt_off, off, dis, sav, lim_ed, vinyl,
                    box_set, deluxe, exclusive, new, blow, sale,  steal, garage, surprise, value)

mod_ctr_all = lm(ctr ~., camp_vars_ctr)
summary(mod_ctr_all)

# compiling results
mod_ctr_all_R_squared <- summary(mod_ctr_all)$r.squared 
mod_ctr_all_F_value <-   summary(mod_ctr_all)$f[1]
mod_ctr_all_re <- c(mod_ctr_all_R_squared, mod_ctr_all_F_value)



# model after removing insignificant variables

mod_ctr_1 = lm(ctr ~ day + length + perc_off + amt_off + off + dis + vinyl + new + blow + value + steal + garage + surprise, camp_vars_ctr)
summary(mod_ctr_1)

# compiling results
mod_ctr_1_R_squared <- summary(mod_ctr_1)$r.squared 
mod_ctr_1_F_value <-   summary(mod_ctr_1)$f[1]
mod_ctr_1_res = c(mod_ctr_1_R_squared, mod_ctr_1_F_value)



# model without the insignificant variables

mod_ctr_2 = lm(ctr ~  length + perc_off + amt_off + off + dis  + blow + value + steal + surprise, camp_vars_ctr)
summary(mod_ctr_2)


mod_ctr_2_R_squared <- summary(mod_ctr_2)$r.squared 
mod_ctr_2_F_value <-   summary(mod_ctr_2)$f[1]
mod_ctr_2_res <-  c(mod_ctr_2_R_squared, mod_ctr_2_F_value)

# removing discoung variable that's now insignificant

mod_ctr_3 = lm(ctr ~  length + perc_off + amt_off + off +  blow + value + steal + surprise, camp_vars_ctr)
summary(mod_ctr_3)


mod_ctr_3_R_squared <- summary(mod_ctr_3)$r.squared 
mod_ctr_3_F_value <-   summary(mod_ctr_3)$f[1]
mod_ctr_3_res <-  c(mod_ctr_3_R_squared, mod_ctr_3_F_value)


# Comparing all three Click Through Rate models:
mod_ctr_res = rbind(mod_ctr_all_re, mod_ctr_1_res,mod_ctr_2_res, mod_ctr_3_res)

```


### Comparing results from al 6 models - CTR and OR

```{r model_comp, eval = TRUE, echo = FALSE, message = FALSE }
mod_or_ctr_res <- as.data.frame(rbind(mod_or_res, mod_ctr_res))
names(mod_or_ctr_res) <- c("R_Squared", "F_Statistic")

```


### Best Model:

```{r model_best, eval = TRUE, echo = FALSE, message = FALSE }
summary(mod_ctr_3)

```
### Preliminary Conclusion:

1. The most effective terms to convey a discount are "Blowout",  and "Steal". 
2. The Value -- as proxy for Artist popularity -- has only a very slight impact of Open and Click Through rates
3. Formats, such as Box Set, Deluxe, etc. aren't associated with higher open rates. 

Maybe we should seprate Artist specific campaigns and general campaigns. The former will appeal to a smaller audience, producing lower open and click through rates, but higher conversion. The latter may will apeal to a broader audience and be associtated with higher open and click through rates, but lower conversion. Let's take a quick look:


### Artist specific campaigns vs generic campaigns

```{r}
camp_artist = camp[camp$Artists != "",]
camp_artist_vars_or <- select(camp_artist, or, hour, Day, length, perc_off, amt_off, off, dis, sav, lim_ed, vinyl,
                    box_set, deluxe, exclusive, new, blow, sale, steal, garage, surprise, value)

lm_camp_artist_or = lm(or ~ ., camp_artist)
summary(lm_camp_artist_or)
camp_generic = camp[camp$Value == 0,]
```



